### 2014-10-31 application.tex
Added coordination and language sections.

@@

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
// TREE BUILDING
// continue tree
waiting(A), token(A, All, Next) -o token(A, All, Next).
// start tree
waiting(A), *underline{@priority(A, A, P)}, P > 0.0
   -o token(A, [A], [A]).
// end tree building
token(A, All, Next), length(All) > maxnodes
   -o first-phase(A, All, reverse(All)).
// expand tree
token(A, All, [A | Next])
   -o [collect => L | Side | !edge(A, L, Side),
         0 = count(All, L),
         0 = count(Next, L),
         *underline{priority(A, L, P)}, P > 0.0,
         *underline{cpu-id(A, L, Id1)},
         *underline{cpu-id(A, A, Id2)}, Id1 = Id2 |
         send-token(A, All, Next ++ L)].

>>>>

Randomized and approximation
algorithms can obtain significant benefits from coordination directives because although the final
program results will not be exact, they follow important statistical properties and can be computed faster.
Examples of such programs are PageRank~\cite{Lubachevsky:1986:CAA:4904.4801} and
Loopy Belief Propagation~\cite{Gonzalez+al:aistats09paraml}.

@@

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\{\}]
set-static(A). // all nodes are static.

// TREE BUILDING
// continue tree
waiting(A), token(A, All, Next) -o token(A, All, Next).
// start tree
waiting(A), *underline{@priority(A, A, P)}, P > 0.0
   -o token(A, [A], [A]).
// end tree building
token(A, All, Next), length(All) > maxnodes
   -o first-phase(A, All, reverse(All)).
// expand tree
token(A, All, [A | Next])
   -o [collect => L | Side | !edge(A, L, Side),
         0 = count(All, L),
         0 = count(Next, L),
         *underline{priority(A, L, P)}, P > 0.0,
         *underline{cpu-id(A, L, Id1)},
         *underline{cpu-id(A, A, Id2)}, Id1 = Id2 |
         send-token(A, All, Next ++ L)].

### 2014-10-31 coordination.tex
Added coordination and language sections.

The original Meld was
implemented as an ensemble programming language, targeting modular robotic systems such as
Claytronics~\cite{ashley-rollman-derosa-iros07wksp}. In such systems, there is a natural matching
between computation and processing units, since each robot is represented by a node. This distribution
of data leaves little choice to be made in the division of computation to the various nodes.

On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
since nodes are a program abstraction and part of the program's logic.
We view the set of nodes as a graph data structure where workers will perform work.
A worker is able to process any node, although a node cannot be computed by more than one worker
at the same time. This disallows the manipulation of a node by multiple workers.

Since LM uses linear logic, the order in which nodes are scheduled can impact the
performance and even the results of the program. When only using
classical logic, no matter how computation is done, the end result will be the same since the program
is strictly monotonic.
LM introduces the concept of \emph{coordination} that allows the programmer
to write code that changes how the runtime system schedules nodes.

Randomized and approximation
algorithms can obtain significant benefits from coordination directives because although the final
program results will not be exact, they follow important statistical properties and can be computed faster.
Examples of such programs are PageRank~\cite{Lubachevsky:1986:CAA:4904.4801} and
Loopy Belief Propagation~\cite{Gonzalez+al:aistats09paraml}.

@@

We can use action facts to change the order in which nodes are evaluated by adding
\emph{priorities}. Node priority works at the worker level
so that each worker can make a local decision about which node of the graph to run next.
Note that, by default, nodes are picked arbitrarily (a FIFO approach is used).

@@

\begin{description}
   \item[\texttt{type linear action set-priority(node, float)}]: This sets the priority of a node. If the node already has some priority, we only change the priority if the new one is higher priority. The programmer can decide if priorities are to be ordered in ascending or descending order.
   \item[\texttt{type linear action add-priority(node, float)}]: This gets the current node's priority and increases or decreases it.
   \item[\texttt{type linear action schedule-next(node)}]: The work will fetch the highest priority node's priority $P$ from its set of nodes and set the action's argument node's priority as $P + 1.0$. If using the priorities
   in ascending order, we pick the lowest priority and subtract $1.0$.
   \item[\texttt{type linear action unset-priority(node)}]: Removes the priority, if any, of a given node.
   \item[\texttt{type linear action stop-program(node)}]: Immediately stops the execution of the whole program.
\end{description}

When the highest priority node is picked up for execution, its priority is reset to 0 (the default priority value). This means that
the programmer must set the node's priority again if he wants to prioritize that node.

We intend to add more action facts in the near future. For example, we want the programmer to be able to place specific nodes in workers. This will permit good use of
memory locality by forcing certain computations to be performed in the same worker.

Sensing facts provide information about node placement and node priority. We can use those facts
to express coordination policies. LM provides the following two
sensing facts:

\begin{description}
   \item[\texttt{type linear cpu-id(node, node, int)}]: The third argument indicates the worker's ID where the node of the second argument is currently running.
   \item[\texttt{type linear priority(node, node, float)}]: The third argument is the current priority of the node in the second argument.
\end{description}

Note that when sensing facts are consumed, they are re-derived automatically, except if the programmer explicitly tells the compiler otherwise.

\subsubsection{Global Directives}

@@

>>>>

Since LM uses linear logic and supports deletion of facts, the order in which
nodes are scheduled can impact the performance and even the results of the
program.

The SSSP program present in Fig.~\ref{code:shortest_path_program} is concise and
declarative but its performance may depend in the order in which nodes are
executed. If nodes with greater distances are prioritized over other nodes, the
program will generate more \texttt{relax} facts since it will take longer to
reach the shortest distances. From Fig.~\ref{fig:shortest_path_program}, it is
clear that the best computational ordering is the following: \texttt{@1},
\texttt{@3}, \texttt{@2} and then \texttt{@4}, where only 4 \texttt{relax}
facts are generated. If we had decided to process nodes in order
\texttt{@1}, \texttt{@2}, \texttt{@4}, \texttt{@3}, \texttt{@4},
\texttt{@2}, then 6 \texttt{relax} facts would have been generated.
Therefore, the optimal solution for SSSP is to schedule the node with the
shortest distance, which is essentially the Dijkstra shortest path
algorithm~\cite{Dijkstra}. Note how it is possible to change the nature of
the algorithm by simply changing the order of node computation, but still
retain the declarative nature of the program.

We introduce the concept of \emph{coordination facts} that allow the programmer
to change how the run time schedules nodes and partitions the nodes among
threads of execution. Coordination facts can be used in the body or head of the
rules to allow the programmer to reason and change how execution is to be done.
In this fashion, we keep the language declaratively because we reason logically
about the state of execution, without the need to introduce extra-logical
operators into the language that would reduce problems when attempting to prove
properties about programs.

There are two kinds of coordination facts. The first kind of coordination facts are
called \emph{sensing facts} and are used to sense information about data
placement (where is the node being executed) or scheduled (what is the priority
of the node). The second kind of coordination facts are \emph{action facts} that,
when derived, are consumed to apply a coordination operation during execution.
These logical rules that use coordination are implemented efficiently in the run
time system and are hidden from the programmer.

@@

We can use action facts to change the order in which nodes are evaluated by adding
\emph{priorities}. Node priority works at the thread level
so that each thread can make a local decision about which node of the graph to run next.
Note that, by default, nodes are picked arbitrarily (a FIFO approach is used).

We have two kinds of priorities: a \emph{temporary priority} and a \emph{default
priority}. A default priority is used indefinitely, while a temporary priority
is only used once, so that when node with a temporary priority $P$ is picked to
run, the priority will default back to $D$, the default priority. Initially,
all nodes have a default priority of $0$.

@@

\begin{itemize}
   \item \texttt{type linear set-priority(node, float)}: This sets the
   temporary priority of a node. If the node already has a priority, we only
   change the priority if the new one is higher priority. The programmer can
   decide if priorities are to be ordered in ascending or descending order.
   \item \texttt{type linear action add-priority(node, float)}: Increments,
   temporarily, the current priority of the node.
   \item\texttt{type linear schedule-next(node)}: The work will fetch the
   highest priority node's priority $P$ from its set of nodes and set the
   action's argument node's priority as $P + 1.0$. If using the priorities in
   ascending order, we pick the lowest priority and subtract $1.0$.
   \item\texttt{type linear set-default-priority(node, float)}: Sets the default
   priority of the node.
   \item \texttt{type linear action stop-program(node)}: Immediately stops the
   execution of the whole program.
\end{itemize}

LM provides the sensing fact \texttt{type linear priority(node, node, float)},
where the second argument is the target node and the third argument is the
node priority. Sensing facts are only used in the body of the rules in order
to fetch information from the runtime system.
Note that when sensing facts are consumed, they are re-derived automatically,
except if the programmer explicitly tells the compiler otherwise. 
Logically, \texttt{set-priority} and \texttt{set-default-priority} update the
value of \texttt{priority} facts, but this done seamlessly by the runtime
system.

\subsubsection{Partitioning facts}

We provide several coordination facts for dealing with node partitioning among
the available threads executing the program. In terms of action
facts, we have the following:

\begin{itemize}
   \item \texttt{type linear set-cpu(node, int)}: Moves the node to a
   thread of execution.
   \item \texttt{type linear set-affinity(node, node)}: Places the first node in
   the same thread as the second node.
   \item \texttt{type linear set-moving(node)}: Allows the node to move freely
   between threads.
   \item \texttt{type linear set-static(node)}: Forces the node to stay in the
   same thread indefinitely.
\end{itemize}

For sensing facts, we have the following set of coordination facts:

\begin{itemize}
   \item \texttt{type linear cpu-id(node, node, int)}: The third argument
   indicates the thread where the node of the second argument is currently running.
   \item \texttt{type linear moving(node)}: Fact available if the node in the
   second argument is allowed to move between threads.
   \item \texttt{type linear static(node, node)}: Fact available if the node in
   the second argument is not allowed to move between threads.
\end{itemize}

\iffalse
\subsubsection{Global Directives}

@@

\fi

### 2014-10-31 implementation.tex
Added coordination and language sections.

>>>>

\subsubsection{Node collection}

Whenever a node has an empty database and no references from other nodes, it is
deleted from the graph. The virtual machine detects such cases by keeping a
count of fact references to every node. Once the count drops to zero and the
database is empty, the node is immediatelly deleted in order to reduce memory
usage.

### 2014-10-31 intro.tex
Added coordination and language sections.

Writing parallel programs in sequential languages is hard because manipulating
shared state using multiple threads of execution may result in data race
conditions. Such issues are handled with low level constructs such as locks and
condition variables, requiring a fair amount of effort to get right.
Declarative programming has been hailed as a solution to this issue, since the
problem of implementing the details of parallelism is moved away from the programmer to the
compiler and runtime of the language. The programmer writes code
without having to deal with parallel programming constructs and the compiler
automatically parallelizes the program in order to take advantage of
multiple threads of execution.
The programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although more general declarative languages have yet to be as successful, the
future looks promising for this particular approach.

We argue, however, that declarative programming leaves little to no programmer control
over how execution is scheduled or how data is laid out. This introduces some
performance issues since a runtime system may be able to parallelize the program
up to a point, using a general algorithm, but lacks some specific information that would make
execution better in terms of execution time, memory usage and scalability.

In this paper, we extend Linear Meld (LM), a declarative language by
\cite{cruz-iclp14}, with coordination facts that give programmer control over
scheduling and data placement. LM is a linear logic programming language
intended for programs over graph structures and with
support for structured manipulation of mutable state using 
linear logic~\cite{girard-87}. In LM, the nodes of the graph perform computation
through derivation of rules and communicate with other nodes when logical facts
are derived.

@@

To the best of our knowledge, this is the first time that a declarative language allows so much
control over execution without a significant change in the language itself.
Our main contributions of this paper are then two-fold: a novel way to coordinate
declarative programs through user control with or without modifications to the
original algorithm, introducing measurable performance improvements; several
examples that take simple declarative algorithms and make them more
efficient through the introduction of coordination facts in the algorithmic
logic.

>>>>

Writing parallel programs in sequential languages is hard because manipulating
shared state using multiple threads may result in data race
conditions. Such issues are handled with low level constructs such as locks and
condition variables, requiring a fair amount of effort to get right.
Declarative programming has been hailed as a solution to this issue, since the
problem of implementing the details of parallelism is moved away from the programmer to the
compiler and runtime of the language. The programmer writes code
without having to deal with parallel programming constructs and the compiler
automatically parallelizes the program in order to take advantage of
multiple threads of execution.
The programming paradigm has been adopted with huge success in domain specific
languages such as SQL and MapReduce~\cite{Dean:2008:MSD:1327452.1327492}.
Although more general declarative languages have yet to be as successful, the
future looks promising for this particular approach.

We argue, however, that declarative programming leaves little to no programmer control
over how execution is scheduled or how data is laid out. This introduces some
performance issues because even if the runtime system is able to reasonably
parallelize the program using a general algorithm, there is a lack of specific
information about the program that a compiler does not easily understand. In
turn, such information would make execution better in terms of run time time,
memory usage and scalability.

In this paper, we extend Linear Meld (LM), a declarative language by
\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer control over
scheduling and data placement. LM is a linear logic programming language
intended for programs over graph structures and with
support for structured manipulation of mutable state using 
linear logic~\cite{girard-87}. In LM, the nodes of the graph perform computation
through derivation of rules and communicate with other nodes when logical facts
are derived.

@@

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution without a significant change in the language itself and
without using non-declarative operations. This proves specially important since
programs can proven to be correct even in the presence of coordination facts.
Our other two main contributions are as follows: a novel way to coordinate
declarative programs through user control that introduce measurable performance
improvements; several application where we add programmer control to declarative
programs in order to make them perform better.

### 2014-10-31 language.tex
Added coordination and language sections.

give an overview of the language

we already have the ICLP paper that describes the language, so we can be brief
here. \cite{cruz-iclp14}

>>>>

Linear Meld (LM) is a forward-chaining linear logic programming
language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
Programs are written as a set of rules of the form
\texttt{a(X), b(Y) -o c(Z)} and are read as following: if \texttt{a(X), b(Y)} (the body of
the rule) is satisfied using some facts \texttt{a(X)} and \texttt{b(Y)} then
\texttt{c(Z)} (the head of the rule) is true.
The facts used during rule derivation are placed in a \emph{database of facts}.
Logical facts are composed of a predicate and a tuple of concrete values,
representing the arguments.
Since LM uses linear logic as its foundation, rule derivations
may delete facts used in the body of the rule. Program execution starts by
adding the \emph{axioms} (the initial facts) of the program to the database.
Next, the rules are recursively applied and the database is updated with new and
deleted facts. When nore more rules are applicable, the program terminates.

LM has been designed for writing graph-based programs. LM accomplishes this by
partitioning the logical facts by the first argument (the \emph{node}) and by
restricting the body of every rule to only refer to the same first argument.
This allows rules to be executed using only facts of a single node. Although the
body of rules is restricted, the head of the rule may refer to any other node as
long as the node variable is refered somewhere in the body of the rule. This allows
derivations of facts in other nodes. The whole database is then composed of smaller
databases (the nodes) and rule derivations allows smaller databases to derive
facts in other databases, creating edges in the graph.

In order to fully understand how the language works, we present a very simple
algorithm: the single source shortest path program~(SSSP). Later in the paper, we
add coordination facts to improve the execution of the program.

The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
with the declaration of the predicates. Predicates specify the kinds of facts
used in the program. The first predicate, \texttt{edge}, specifies edges between
nodes of the graph, where the third argument represents the weight of the edge.
The \texttt{route} modifier informs the compiler about the structure of the
program graph. The predicates \texttt{shortest} and \texttt{relax} are specified
as linear facts and thus are deleted when used in rules.
The idea of the algorithm is to compute the shortest distance from node
\texttt{@1} to all other nodes in the graph. Every node has a \texttt{shortest}
fact that is improved with new \texttt{relax} facts.
Lines 5-9 declare the axioms of the program: \texttt{edge} facts describe the
graph; \texttt{shortest(A, +00, [])} is the initial shortest distance (infinity)
for all nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm with the
initial distance from \texttt{@1} to \texttt{@1}.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left]
type route edge(node, node, int).
type linear shortest(node, int, list int).
type linear relax(node, int, list int).

!edge(@1, @2, 3). !edge(@1, @3, 1).
!edge(@3, @2, 1). !edge(@3, @4, 5).
!edge(@2, @4, 1).
shortest(A, +00, []).
relax(@1, 0, [@1]).

relax(A, D1, P1), shortest(A, D2, P2),
D1 < D2
   -o shortest(A, D1, P1),
      {B, W | !edge(A, B, W) |
         relax(B, D1 + W, P1 ++ [B])}.

relax(A, D1, P1), shortest(A, D2, P2),
D1 >= D2
   -o path(A, D2, P2).
\end{Verbatim}
  \caption{Single Source Shortest Path program code.}
  \label{code:shortest_path_program}
\end{figure}
\normalsize

The first rule of the program (lines 11-15) replaces the current path in
\texttt{shortest} with a shorter one in \texttt{relax}. The rule deletes both
\texttt{relax} and \texttt{shortest} facts and derives a new \texttt{shortest}
fact. In lines 14-15, we have a \emph{comprehension} where the program iterates
over the edges of node \texttt{A} and derives a new \texttt{relax} fact at
\texttt{B} with the new shorter distance plus the weight of the edge. For
instance, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
applying the same rule but at node \texttt{2}.

\begin{figure*}[ht]
\begin{center}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest2}}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest3}}
  \subfloat[]{\includegraphics[width=0.3\textwidth]{figures/shortest8}}
\end{center}
\caption{Graphical representation of the SSSP program. Figure (a) represents the
   program after propagating initial distance at node \texttt{@1}, followed by
   Figure (b) where the first rule is applied in node \texttt{@3}. Figure (c)
   represents the state of the final program, where all the shortest paths
   have been computed.}
\label{fig:shortest_path_program}
\end{figure*}

The second rule of the program (lines 17-19) deals with the case where the new
distance in \texttt{relax} is not better than the current one, so it is thrown
away and the current distance is kept.

There many opportunities for concurrency in the SSSP program. For instance,
after applying rule 1 in Fig.~\ref{fig:shortest_path_program}~(a), it is
possible to either apply rules in either node \texttt{@2} or node
\texttt{@3}. In our implementation, we partition the graph into subgraphs that
are processed by multiple threads of execution. Eventually, it is no longer to
possible to apply rules and the final result present in
Fig.~\ref{fig:shortest_path_program}~(c) is achieved.

%On the other hand, LM has no natural matching of data and computation to workers (processes, threads),
%since nodes are a program abstraction and part of the program's logic.
%We view the set of nodes as a graph data structure where workers will perform work.
%A worker is able to process any node, although a node cannot be computed by more than one worker
%at the same time. This disallows the manipulation of a node by multiple workers.
